### 2013-08-26 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding pauls changes to chapter 4

\subsection{Inexact Newton Methods}
\label{subsec:inexact_newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted as initially described by Dembo and others
\cite{dembo_inexact_1982}. For common sparse nonlinear systems, which
in turn yield a sparse Jacobian matrix, this situation occurs when
conventional iterative methods are applied. In their definition, Dembo
formulated inexact methods such that they are independent of the
linear method used to solve for the Newton correction and therefore
are amenable to use with any linear solver. Furthermore, they bind the
convergence of the outer nonlinear iteration to the inner linear
iteration such that:
\begin{equation}
  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
  ||\ve{F}(\ve{u}^k)||\:,
  \label{eq:inexact_newton_forcing}
\end{equation}
where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
  that the residual generated by the linear solver is bound by the
  nonlinear residual and how tightly it is bound is defined by the
  forcing term. This is useful in that we can vary how tightly coupled
  the convergence of the linear iterations used to generate the Newton
  correction is to the nonlinear iteration by relaxing or tightening
  the convergence properties on the linear iterative method. As a
  result, strategies for determining the forcing term can vary
  depending on the problem type and can greatly affect the convergence
  of the method or even prohibit convergence
  \cite{eisenstat_choosing_1996}. In addition, \textit{globalization
    methods} may be used to modify the Newton correction in a more
  desirable direction such that nonlinear convergence properties can
  be improved when the initial guess for $\ve{u}$ is poor
  \cite{pawlowski_globalization_2006}.

@@

At first glance, FANM is simply a variant of an inexact Newton method
with a specific requirement of a fully formed Jacobian. However, FANM
has the potential to provide value in several areas. First, if
iterative or parallel performance of MCSA can be demonstrated to
better than that of GMRES or similar subspace methods for asymmetric
systems, FANM may outperform a Newton-Krylov method. Second, for
situations where memory is a constraint, especially in situations
where the Jacobian is generated often if not at every Newton
iteration, not building a subspace for ill-conditioned problems will
alleviate the associated cost. However, until the preconditioning
issues associated with MCSA as presented in the previous chapter using
the $SP_N$ equations can be resolved, this hypothesis can
unfortunately not be researched. Explicit algebraic preconditioning
used in conventional solution methods will almost certainly be
required to reduce the spectral radius of most Jacobian operators for
MCSA to apply and therefore the current preconditioning strategy will
eliminate all potential memory benefits of FANM.

@@

For each of the benchmark problems, we will present the results of
computations using FANM and compare them to results using a
Newton-Krylov method in order to verify the correctness of FANM and
its applicability to fluid problems. Given the difficulty of these
benchmark problems, the linear solve at each Newton step was
preconditioned with an algebraic multigrid method\footnote{The ML
  algebraic multigrid library in the Trilinos scientific computing
  libraries \cite{heroux_overview_2005} was used for all problems.}, a
common technique for preconditioning these equations
\cite{ghia_high-re_1982,evans_enhanced_2007}. The preconditioner is
reconstructed at every Newton step for the new Jacobian and used on
the right as outlined in
\S~\ref{subsubsec:general_mcsa_preconditioning}. Without multigrid
preconditioning, the residual in MCSA was observed to
stagnate. Typically this happens because the Richardson iteration and
subsequently the Neumann series will quickly damp the higher order
error modes on the scale of the computational grid used to solve the
problem but will reduce lower order error modes at a considerably
lower rate. In terms of Monte Carlo, this stagnation manifests itself
in histories that simply do no terminate and therefore no solution is
acquired. For many of the problems presented here, it is likely that a
more physics-based preconditioning strategy such as those presented in
\cite{evans_development_2006,evans_enhanced_2007} will yield better
results for more difficult problems. However, in this work we will
restrict our attention to multigrid for simplicity.

@@

Figure~\ref{fig:ra1e5_convergence} gives the nonlinear residual
behavior for the \sn{1}{5} Rayleigh number case that required the
extra FANM iteration. We see that the extra iteration comes from
slightly slower quadratic convergence with FANM for this particular
case. The difference in convergence of the nonlinear problem comes
from the fact that the same multigrid preconditioner parameters are
used for this case as for the \sn{1}{3} and \sn{1}{4} Rayleigh number
cases for the linear solutions. Using the same preconditioner for the
harder \sn{1}{5} case means that the problem is still growing in
difficulty even though it is preconditioned in an identical
manner. This means that the GMRES solver at each Newton step is
converging the linear residual to a smaller tolerance than the MCSA
solver as the forcing terms adapt to the convergence of the method
over the course of nonlinear iterations. For this problem, tighter
convergence of the linear model at later Newton steps also improved
convergence of nonlinear model, resulting in the extra FANM iteration.

@@

In addition to nonlinear iterative performance, performance of the
inner linear solves is also of interest. Here, we seek to maintain a
minimum number of total linear solver iterations over the course of
all nonlinear iterations in order to limit matrix-vector multiplies
and other costly
operations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES, FANM, and Richardson iterations required to
converge the Newton method at each Rayleigh number used. For all
cases, fewer MCSA iterations were required than GMRES iterations to
converge the problem even with both solvers preconditioned with the
same multigrid package and the same parameters. For the \sn{1}{6}
Rayleigh number case where a constant forcing term of \sn{1}{-4} was
kept, this difference is most drastic with FANM requiring only 64\% of
the linear iterations required by the Newton-Krylov scheme with the
linear solver per nonlinear iteration given for this case in
Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5} Rayleigh
number case that required more Newton iterations required 5 fewer MCSA
iterations to converge than GMRES iterations. In all cases, MCSA
improved on the iterative performance of using only the Richardson
iteration through the residual Monte Carlo acceleration. If the
performance of the MCSA implementation used here were to be
potentially optimized in the future, this improved iterative
performance could also mean improved time to solution.

@@

This result can be explained by the fact that as given by
Table~\ref{tab:driven_mcsa_parameters}, although the preconditioner
parameters were the same for both calculations, the MCSA iteration was
pushed harder by doubling the number of histories used for the
computation at each nonlinear iteration. These extra histories
provided better convergence of the linear model at each iteration and
did not result in oversolving, thereby increasing the convergence rate
of the nonlinear model as well. This result is significant in that it
shows increasing the number of histories used for each MCSA iteration
can improve the performance of both linear and nonlinear convergence.

@@

As an additional means of analysis for the linear solver iteration
results, we will look at a case where MCSA performed poorly compared
to GMRES. Consider the linear iterations as a function of nonlinear
iteration given in Figure~\ref{fig:driven_re500_iters} and the forcing
term as a function of nonlinear iteration given in
Figure~\ref{fig:driven_re500_forcing} for the 500 Reynolds number
case. At the later Newton iterations where the region of fast
quadratic convergence is reached, we see that MCSA required many more
iterations to converge than GMRES, accounting for the majority of the
extra iterations. However, we see in the forcing term data that
forcing terms at later Newton iterations for MCSA were much smaller
than those for GMRES. This means that MCSA was converged to a tighter
tolerance at these iterations and therefore required the extra
iterations.

@@

The backward step problem exhibited different performance
characteristics than the other two benchmark problems, likely due to
the different geometry and boundary
conditions. Table~\ref{tab:step_nonlinear_iter_comparison} gives the
nonlinear iteration results for this benchmark. At low Reynolds
numbers, FANM converges in fewer nonlinear iterations than the
Newton-Krylov solver. However, as the problem becomes more difficult
to solver at higher rates of flow, the Newton-Krylov solver eventually
converges in fewer nonlinear iterations. We see the same behavior as
well in the linear solver iteration data given by
Table~\ref{tab:step_linear_iter_comparison}. At the 3 lowest Reynolds
numbers, FANM required significantly fewer linear solver iterations to
converge the problem while the ill-conditioned case at a Reynolds
number of 500 is observed to use many more MCSA iterations. For
Reynolds numbers of both 200 and 500, FANM also reduced the number of
Newton steps need when compared to the Newton-Richardson scheme.

@@

Unlike the lid driven cavity benchmark, the extra linear solver
iterations are not due to the smaller forcing terms but are rather
truly a function of the ill-conditioning of the Jacobian matrices
produced at each Newton step. We see this by comparing the results at
Reynolds numbers of 400 where FANM outperformed Newton-Krylov and 500
where Newton-Krylov outperformed FANM. Linear iterations as a function
of nonlinear iteration given in Figure~\ref{fig:step_re400_iters} and
the forcing term as a function of nonlinear iteration given in
Figure~\ref{fig:step_re400_forcing} for the 400 Reynolds number case
and the same data is given in Figure~\ref{fig:step_re500_iters} and
Figure~\ref{fig:step_re500_forcing} for the 500 Reynolds number
case. For both cases, FANM always has larger forcing terms and should
therefore use fewer MCSA iterations at each Newton step to converge
the linear model if the iterative performance is comparable. This is
indeed the case for the 400 Reynolds number case where the linear
model is better conditioned and FANM used fewer total linear solver
iterations to converge the nonlinear residual. However, at a Reynolds
number of 500 FANM has larger forcing terms but requires many more
MCSA iterations to converge the last 2 Newton steps. Therefore, these
extra iterations in this case must come from ill-conditioning.

@@

\begin{itemize}
\item Forward-Automated Newton-MCSA (FANM), a new inexact Newton
  method leveraging Monte Carlo Synthetic Acceleration, has been
  developed
\item The FANM method has been incorporated into the Drekar
  multiphysics production code base developed at Sandia National
  Laboratories
\item The FANM method has been verified to produce the same solutions
  as a production Newton-Krylov method for three difficult benchmark
  problems for the Navier-Stokes equations in different flow regimes
  and geometries
\item The FANM method has better iterative performance than the
  Newton-Krylov method for convection dominated problems, converging
  in fewer linear solver iterations with the same preconditioning for
  high and low Rayleigh numbers
\item The Newton-Krylov method has better iterative performance than
  the FANM method for flow dominated by inertial forces, converging in
  fewer linear solver iterations with the same preconditioning for
  high Reynolds numbers
\item The spectral radius convergence restriction on MCSA was observed
  to be a significant hindrance by preventing solutions to forced flow
  problems at high Reynolds numbers
\item Explicit algebraic preconditioning used to overcome the spectral
  radius restriction of MCSA creates FANM run times
  $O(100)$-$O(1,000)$ slower than the Newton-Krylov solver
\item More Monte Carlo histories at every FANM iteration can reduce
  the number of linear and nonlinear iterations required to converge
  the problem
\end{itemize}

>>>>

\subsection{Inexact Newton Methods}
\label{subsec:inexact_newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted as initially described by Dembo and others
\cite{dembo_inexact_1982}. For common sparse nonlinear systems, which
in turn yield a sparse Jacobian matrix, this situation occurs when
conventional iterative methods are applied. In their definition, Dembo
formulated inexact methods such that they are independent of the
linear method used to solve for the Newton correction and therefore
are amenable to use with any linear solver. Furthermore, they bind the
convergence of the outer nonlinear iteration to the inner linear
iteration such that:
\begin{equation}
  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
  ||\ve{F}(\ve{u}^k)||\:,
  \label{eq:inexact_newton_forcing}
\end{equation}
where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
  nonlinear iteration $k$, the right-hand side is the nonlinear
  residual at nonlinear iteration $k$, and the left-hand side is the
  residual for the linear model at nonlinear iteration
  $k$. Eq~(\ref{eq:inexact_newton_forcing}) then states that the
  residual generated by the linear solver is bound by the nonlinear
  residual and how tightly it is bound is defined by the forcing
  term. This is useful in that we can vary how tightly coupled the
  convergence of the linear iterations used to generate the Newton
  correction is to the nonlinear iteration by relaxing or tightening
  the convergence properties on the linear iterative method. As a
  result, strategies for determining the forcing term can vary
  depending on the problem type and can greatly affect the convergence
  of the method or even prohibit convergence
  \cite{eisenstat_choosing_1996}. In addition, \textit{globalization
    methods} may be used to modify the Newton correction in a more
  desirable direction such that nonlinear convergence properties can
  be improved when the initial guess for $\ve{u}$ is poor
  \cite{pawlowski_globalization_2006}.

@@

Upon preliminary inspection, FANM is simply a variant of an inexact
Newton method with a specific requirement of a fully formed
Jacobian. However, FANM has the potential to provide value in several
areas. First, if iterative or parallel performance of MCSA can be
demonstrated to better than that of GMRES or similar subspace methods
for asymmetric systems, FANM may outperform a Newton-Krylov
method. Second, for situations where memory is a constraint,
especially in situations where the Jacobian is generated often if not
at every Newton iteration, not building a subspace for ill-conditioned
problems will alleviate the associated cost. However, until the
preconditioning issues associated with MCSA as presented in the
previous chapter using the $SP_N$ equations can be resolved, this
hypothesis can unfortunately not be researched. Explicit algebraic
preconditioning used in conventional solution methods will almost
certainly be required to reduce the spectral radius of most Jacobian
operators for MCSA to apply and therefore the current preconditioning
strategy will eliminate all potential memory benefits of FANM.

@@

For each of the benchmark problems, we will present the results of
computations using FANM and compare them to results using a
Newton-Krylov method in order to verify the correctness of FANM and
its applicability to fluid problems. Given the difficulty of these
benchmark problems, the linear solve at each Newton step was
preconditioned with an algebraic multigrid method\footnote{The ML
  algebraic multigrid library in the Trilinos scientific computing
  libraries \cite{heroux_overview_2005} was used for all problems.}, a
common technique for preconditioning these equations
\cite{ghia_high-re_1982,evans_enhanced_2007}. The preconditioner is
reconstructed at every Newton step for the new Jacobian and used on
the right as outlined in
\S~\ref{subsubsec:general_mcsa_preconditioning}. Without multigrid
preconditioning, the residual in MCSA was observed to
stagnate. Typically this happens because the Richardson iteration and
subsequently the Neumann series will quickly dampen the higher order
error modes on the scale of the computational grid used to solve the
problem but will reduce lower order error modes at a considerably
lower rate. In terms of Monte Carlo, this stagnation manifests itself
in histories that simply do no terminate and therefore no solution is
acquired. For many of the problems presented here, it is likely that a
more physics-based preconditioning strategy such as those presented in
\cite{evans_development_2006,evans_enhanced_2007} will yield better
results for more difficult problems. However, in this work we will
restrict our attention to multigrid for simplicity.

@@

Figure~\ref{fig:ra1e5_convergence} gives the nonlinear residual
behavior for the \sn{1}{5} Rayleigh number case that required the
extra FANM iteration. We see that the extra iteration comes from
slightly slower quadratic convergence with FANM for this particular
case. The difference in convergence of the nonlinear problem comes
from the fact that the same multigrid preconditioner parameters are
used for this case as for the \sn{1}{3} and \sn{1}{4} Rayleigh number
cases for the linear solutions. Using the same preconditioner for the
harder \sn{1}{5} case means that the problem is still growing in
difficulty even though it is preconditioned in an identical
manner. This means that GMRES at each Newton step is converging the
linear residual to a smaller tolerance than MCSA as the forcing terms
adapt to the convergence of the method over the course of nonlinear
iterations. For this problem, tighter convergence of the linear model
at later Newton steps also improved convergence of nonlinear model,
resulting in the extra FANM iteration.

@@

In addition to nonlinear iterative performance, performance of the
inner linear solves is also of interest. Here, we seek to minimize the
minimum number of total linear solver iterations over the course of
all nonlinear iterations in order to limit matrix-vector multiplies
and other costly
operations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES, FANM, and Richardson iterations required to
converge the Newton method at each Rayleigh number used. For all
cases, fewer MCSA iterations were required than GMRES iterations to
converge the problem even with both solvers preconditioned with the
same multigrid package and the same parameters. For the \sn{1}{6}
Rayleigh number case where a constant forcing term of \sn{1}{-4} was
kept, this difference is most drastic with FANM requiring only 64\% of
the linear iterations required by the Newton-Krylov scheme with the
linear solver per nonlinear iteration given for this case in
Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5} Rayleigh
number case that required more Newton iterations required 5 fewer MCSA
iterations to converge than GMRES iterations. In all cases, MCSA
improved on the iterative performance of using only the Richardson
iteration through the residual Monte Carlo acceleration. If the
performance of the MCSA implementation used here were to be
potentially optimized in the future, this improved iterative
performance could also mean improved time to solution.

@@

This result can be explained by the fact that as given by
Table~\ref{tab:driven_mcsa_parameters}, although the preconditioner
parameters were the same for both calculations, the MCSA iteration was
pushed harder at a Reynolds number of 700 by doubling the number of
histories used for the computation at each nonlinear iteration. These
extra histories provided better convergence of the linear model at
each iteration and did not result in over-solving, thereby increasing
the convergence rate of the nonlinear model as well. This result is
significant in that it shows increasing the number of histories used
for each MCSA iteration can improve the performance of both linear and
nonlinear convergence. However, further research is required to study
the relationship between the acceleration parameters and the nonlinear
convergence to confirm this for a broader range of problems.

@@

As an additional means of analysis for the linear solver iteration
results, we will look at a case where MCSA performed poorly compared
to GMRES. Consider the linear iterations as a function of nonlinear
iteration given in Figure~\ref{fig:driven_re500_iters} and the forcing
term as a function of nonlinear iteration given in
Figure~\ref{fig:driven_re500_forcing} for the 500 Reynolds number
case. At the later Newton iterations where the region of fast
quadratic convergence is reached, we see that MCSA required many more
iterations to converge than GMRES, accounting for the majority of the
extra iterations. However, we see in the forcing term data that
forcing terms at later Newton iterations for MCSA were much smaller
than those for GMRES. Per Eq~(\ref{eq:newton_stopping_criteria}), this
means that MCSA was converged to a smaller tolerance at these
nonlinear iterations than GMRES and therefore required extra linear
iterations to meet the convergence criteria.

@@

The backward step problem exhibited different performance
characteristics than the other two benchmark problems, likely due to
the different geometry and boundary
conditions. Table~\ref{tab:step_nonlinear_iter_comparison} gives the
nonlinear iteration results for this benchmark. At low Reynolds
numbers, FANM converges in fewer nonlinear iterations than the
Newton-Krylov solver. However, as the problem becomes more difficult
to solver at higher rates of flow, the Newton-Krylov solver eventually
converges in fewer nonlinear iterations. We see the same behavior as
well in the linear solver iteration data given by
Table~\ref{tab:step_linear_iter_comparison}. At the 3 lowest Reynolds
numbers, FANM required significantly fewer linear solver iterations to
converge the problem while the ill-conditioned case at a Reynolds
number of 500 is observed to use many more MCSA iterations. For
Reynolds numbers of 200, 300, and 500, FANM also reduced the number of
Newton steps need when compared to the Newton-Richardson scheme.

@@

Unlike the lid driven cavity benchmark, the extra linear solver
iterations are not due to the smaller forcing terms but are rather
truly a function of the ill-conditioning of the Jacobian matrices
produced at each Newton step. We see this by comparing the results at
Reynolds numbers of 400 where FANM outperformed Newton-Krylov and 500
where Newton-Krylov outperformed FANM. Linear iterations as a function
of nonlinear iteration given in Figure~\ref{fig:step_re400_iters} and
the forcing term as a function of nonlinear iteration given in
Figure~\ref{fig:step_re400_forcing} for the 400 Reynolds number case
and the same data is given in Figure~\ref{fig:step_re500_iters} and
Figure~\ref{fig:step_re500_forcing} for the 500 Reynolds number
case. For both cases, FANM always has larger forcing terms and should
therefore use fewer MCSA iterations at each Newton step to converge
the linear model if the iterative performance is comparable. This is
indeed the case for the 400 Reynolds number case where FANM used fewer
total linear solver iterations to converge the nonlinear
residual. However, at a Reynolds number of 500 FANM has larger forcing
terms but requires many more MCSA iterations to converge the last 2
Newton steps. Therefore, these extra iterations in this case must come
from ill-conditioning.

@@

\begin{itemize}
\item Forward-Automated Newton-MCSA (FANM), a new inexact Newton
  method leveraging Monte Carlo Synthetic Acceleration, has been
  developed
\item The FANM method has been incorporated into the Drekar
  multiphysics production code base developed at Sandia National
  Laboratories
\item The FANM method has been verified to produce the same solutions
  as a production Newton-Krylov method for three difficult benchmark
  problems for the Navier-Stokes equations in different flow regimes
  and geometries
\item The FANM method has better iterative performance than the
  Newton-Krylov method for convection dominated problems, converging
  in fewer linear solver iterations with the same preconditioning for
  high and low Rayleigh numbers
\item The Newton-Krylov method has better iterative performance than
  the FANM method for flow dominated by inertial forces, converging in
  fewer linear solver iterations with the same preconditioning for
  high Reynolds numbers
\item Using more Monte Carlo histories at every FANM iteration was
  observed to reduce the number nonlinear iterations required to
  converge the lid-driven cavity problem at high Reynolds numbers
\item The spectral radius convergence restriction on MCSA was observed
  to be a significant hindrance by preventing solutions to forced flow
  problems at high Reynolds numbers
\item Explicit algebraic preconditioning used to overcome the spectral
  radius restriction of MCSA creates FANM run times
  $O(100)$-$O(1,000)$ slower than the Newton-Krylov solver
\end{itemize}
